---
title: "Evaluation of machine learning models for usage-based insurance"
author: "Davide Gentile"
description: "This project leveraged fleet telematics data to inform usage-based insurance models for corporate vehicle fleets. We compared machine learning algorithms to predict collision risk in real-time driving behavior data from 3,854 corporate vehicle drivers. The analysis identified key factors that influenced collision involvement, such as driving time, trip frequency, and rapid speed changes. Fleet rental companies can use these insights to adjust insurance rates based on risky driver behaviors. This project contributes to the development of more accurate risk models and improved operational strategies for commercial fleet services."
date: "2024-08-24"
execute:
  warning: false
---

This analysis takes the best from three analysis on the Penguin data.

Showing mostly a PCA, and Simpson's paradox.

```{r}
library(palmerpenguins)
library(dplyr)
library(tidyr)
data("penguins")
str(penguins)
```

https://allisonhorst.github.io/palmerpenguins/

```{r}
penguins %>% 
  group_by(species) %>% 
  summarize(across(where(is.numeric), mean, na.rm = TRUE))
```

For this example:

-   shorten variable names (remove units) to simplify variable labels,

-   create factors for character variables (needed for MANOVA), and

-   remove NA observations (causes problems with PCA)

```{r}
peng <- penguins %>%
    dplyr::rename(
         bill_length = bill_length_mm, 
         bill_depth = bill_depth_mm, 
         flipper_length = flipper_length_mm, 
         body_mass = body_mass_g)
peng <- peng %>% drop_na()
peng
```

```{r}
library(car)
library(ggbiplot)
library(GGally)
```

```{r}
scatterplotMatrix(~ bill_length + bill_depth + flipper_length + body_mass | species,
                  data=peng,
                  ellipse=list(levels=0.68),
                  col = scales::hue_pal()(3),
                  legend=list(coords="bottomright"))
```

 

```{r}
ggpairs(peng, mapping = aes(color = species), 
        columns = c("bill_length", "bill_depth", 
                    "flipper_length", "body_mass",
                    "island", "sex"))
```

```{r}
peng.pca <- prcomp (~ bill_length + bill_depth + flipper_length + body_mass,
                    data=peng,
                    scale. = TRUE)

peng.pca
```

```{r}
screeplot(peng.pca, type = "line", lwd=3, cex=3, 
        main="Variances of PCA Components")
```

```{r}
ggbiplot(peng.pca, obs.scale = 1, var.scale = 1,
         groups = peng$species, 
         ellipse = TRUE, circle = TRUE) +
  scale_color_discrete(name = 'Penguin Species') +
  theme_minimal() +
  theme(legend.direction = 'horizontal', legend.position = 'top') 
```

From this, we can see:

-   These two principal components account for 68.6 + 19.5 = 88.1 % of the total variance of these four size variables.

-   PC1 is largely determined by flipper length and body mass. We can interpret this as an overall measure of **penguin size**.

-   On this dimension, Gentoos are the largest, by quite a lot, compared with Adelie and Chinstrap.

-   PC2 is mainly determined by variation in the two beak variables: bill length and depth. Chinstrap are lower than the other two species on bill length and depth, but bill length further distinguishes the Gentoos. A penguin biologist could almost certainly provide an explanation, but I'll call this **beak shape**.

```{r}

# Scatterplot example 2: penguin bill length versus bill depth
ggplot(data = peng, aes(x = bill_length, y = bill_depth)) +
  geom_point(aes(color = species, 
                 shape = species),
             size = 2)  +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme_minimal()
```
